Real-space computation of $E$/$B$-mode maps I: Formalism, Compact Kernels, and Polarized Filaments

### 2018-07-31 articles/formalism/sections/discussion.tex
removed text change highlights and added acknowledgements
Added some clarifications based on Mathieu comments.

These real space operators offer a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts. While the azimuthal part of the operators is primarily responsible for handling the spin decomposition, the radial weights determine the non-local dependence of the resulting fields on the original fields.  Only the radial part of the kernel depends on the band limit. 
The radial parts of the operator kernels are roughly self-similar in the sense that the radial kernels evaluated with some band limit are related to other radial kernels (evaluated with a different band limit) by an approximate rescaling of the function. We use this property to define a non-locality parameter $\beta_0$ as the angular distance at which the amplitude of the radial kernel falls below one percent of its maxima and it  is a function of the maximum multipole $\ell_{\rm max}$ available for the analysis. We empirically show that the non-locality parameter is approximately given by: $\beta_0 = \mathrm{min}(180^{\circ}, 180^{\circ} \frac{\ell_{0}}{\ell_{\rm max}})$ with $\ell_{0}=22$ for the operator that converts Stokes Q/U to scalars E/B and vice versa (see \fig{fig:rad_ker_decay}). An analysis in  \cite{Zaldarriaga2001a} treated real space $E/B$ operators in the flat sky.  It did not explicitly derive the radial part of the kernel, but argued on geometric grounds that it should fall with angular separation as $\beta^{-2}$.  We find that this agrees with the average behavior of radial functions after averaging oscillations and note that some such averaging always takes place in practice due to pixelization of the signal.  However, for precision reconstruction of $E/B$ on the sphere, the oscillations must be taken into account.

Our careful study of the real space operators show that they can be expressed either as Green's functions or as convolving beam functions.  The convolution interpretation is not a totally new concept.  It guides the discussion in \cite{Zaldarriaga2001a} and closely relates to the popular spokes and pinwheel descriptions of the $E/B$ modes.  However, the radiation/Green's function interpretation of the operators is a new one and is discussed here for the first time in detail on the sphere. These two different interpretations of the operators emerge from the expression of the kernels in terms of the forward or inverse rotation Euler angles.  The mathematical forms of the Green's function and convolution kernels swap roles (and are conjugated) when transforming back to $Q/U$ from $E/B$.

@@

Since the real space operators let us tune the locality of $E/B$ maps, this can be potentially exploited to eliminate foreground contamination from distant parts of the sky. Such applications can be difficult to implement using conventional harmonic space methods.  For instance, the real space operators can be defined such that the locality of their radial kernels is varied on different portions of the sky, dictated by say the foreground morphology, resulting in some modified scalar $E'/B'$ maps.  While this idea seems interesting, the usefulness of this implementation will depend on whether the standard $E/B$ mode spectra are easily recoverable in this fashion.  We will explore some of these possible analysis directions in the future papers of this series.

>>>>

These real space operators offer a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts. While the azimuthal part of the operators is primarily responsible for handling the spin decomposition, the radial weights determine the non-local dependence of the resulting fields on the original fields.  Only the radial part of each kernel depends on the band limit. 
The radial parts of the operator kernels are roughly self-similar in the sense that the radial kernels evaluated with some band limit are related to other radial kernels (evaluated with a different band limit) by an approximate rescaling of the function. We use this property to define a non-locality parameter $\beta_0$ as the angular distance at which the amplitude of the radial kernel falls below one percent of its maxima and it  is a function of the maximum multipole $\ell_{\rm max}$ available for the analysis. We empirically show that the non-locality parameter is approximately given by: $\beta_0 = \mathrm{min}(180^{\circ}, 180^{\circ} \frac{\ell_{0}}{\ell_{\rm max}})$ with $\ell_{0}=22$ for the operator that converts Stokes $Q/U$ to scalars $E/B$ and vice versa (see \fig{fig:rad_ker_decay}). An analysis in  \cite{Zaldarriaga2001a} treated real space $E/B$ operators in the flat sky.  It did not explicitly derive the radial part of the kernel, but argued on geometric grounds that it should fall with angular separation as $\beta^{-2}$.  We find that this agrees with the average behavior of radial functions after averaging oscillations and note that some such averaging always takes place in practice due to pixelization of the signal.  However, for precision reconstruction of $E/B$ on the sphere, the oscillations must be taken into account.

Our careful study of the real space operators show that they can be expressed either as Green's functions or as convolving beam functions.  The convolution interpretation is not a totally new concept.  It guides the discussion in \cite{Zaldarriaga2001a,Chiueh2002} and closely relates to the popular spokes and pinwheel descriptions of the $E/B$ modes. However, the radiation/Green's function interpretation of the operators is a new one and is discussed here for the first time in detail on the sphere. These two different interpretations of the operators emerge from the expression of the kernels in terms of the forward or inverse rotation Euler angles.  The mathematical forms of the Green's function and convolution kernels swap roles (and are conjugated) when transforming back to $Q/U$ from $E/B$.

@@

Since the real space operators let us tune the locality of $E/B$ maps, this can be potentially exploited to eliminate foreground contamination from distant parts of the sky. Such applications can be difficult to implement using conventional harmonic space methods.  For instance, the real space operators can be defined such that the locality of their radial kernels is varied on different portions of the sky, dictated by say the foreground morphology, resulting in some modified scalar $E'/B'$ maps.  While this idea seems interesting, the usefulness of this idea will depend on whether the standard $E/B$ mode spectra are easily recoverable in this fashion.  We will explore some of these possible analysis directions in the future papers of this series.

### 2018-07-31 articles/formalism/sections/generalizedeb.tex
Added some clarifications based on Mathieu comments.

%--------------------------------------------------------
\paragraph{Relation to the spin raising $\eth^2$ and lowering $\bar{\eth}^2$ operators.}
Recall that on operating twice with the spin lowering operator on the Stokes charge ${}_{+2}X$ results in filtered version of $E/B$ maps as in \eq{eq:ebdef}. Now note that it is possible to construct a modified real space operator by choosing the harmonic space function to be $g_{\ell} = ({(\ell+2)!/(\ell-2)!})^{1/2}$, resulting in similarly filtered $E/B$ maps as follows:
%
\beq 
[\mathcal{E} + i \mathcal{B}](\hat{n}_e)=- \Delta \Omega\sum_{q=1}^{N_{\rm pix}} \Bigg\lbrace  \left[  \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} P_{\ell}^2(\beta_{qe}) \right] e^{-i2\alpha_{eq}} {}_{+2}X(\hat{n}_{q}) \Bigg\rbrace \,. \label{eq:bl_ebdef_lower} 
\eeq
%
Comparing to \eq{eq:ebdef_lower} makes apparent the following mapping:
%
\beq
\bar{\eth}^2_{\textrm BL} \equiv \Delta \Omega \sum_{q=1}^{N_{\rm pix}} \left[ \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} P_{\ell}^2(\beta_{qe}) \right] e^{-i2\alpha_{eq}}\,,
\eeq
%
where we use the notation $\bar{\eth}^2_{\textrm BL}$ to represent the band limited version of the spin lowering operator, thats relevant in the case of a discretely sampled spin-2 field on the sphere. In the limit of $\ell_{\rm max} \rightarrow \infty$, the $\bar{\eth}^2_{\textrm BL} \rightarrow \bar{\eth}^2$.
The band limited version of the spin raising operator $\eth^2$ is derived by simply taking the conjugate of the above equation.
%--------------------------------------------------------

>>>>

%--------------------------------------------------------
\paragraph{Relation to the spin raising $\eth^2$ and lowering $\bar{\eth}^2$ operators.}
Recall that on operating twice with the spin lowering operator on the Stokes charge ${}_{+2}X$ results in filtered version of $E/B$ maps as in \eq{eq:ebdef}. Now note that it is possible to construct a modified real space operator by choosing the harmonic space function to be $g_{\ell} = ({(\ell+2)!/(\ell-2)!})^{1/2}$, resulting in similarly filtered $E/B$ maps as follows:
%
\beq 
[\mathcal{E} + i \mathcal{B}](\hat{n}_e)=- \Delta \Omega\sum_{q=1}^{N_{\rm pix}} \Bigg\lbrace  \left[  \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} P_{\ell}^2(\beta_{qe}) \right] e^{-i2\alpha_{eq}} {}_{+2}X(\hat{n}_{q}) \Bigg\rbrace \,. \label{eq:bl_ebdef_lower} 
\eeq
%
Comparing to \eq{eq:ebdef_lower} makes apparent the following mapping:
%
\beq
\bar{\eth}^2_{\textrm BL} \equiv \Delta \Omega \sum_{q=1}^{N_{\rm pix}} \left[ \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} P_{\ell}^2(\beta_{qe}) \right] e^{-i2\alpha_{eq}}\,,
\eeq
%
where we use the notation $\bar{\eth}^2_{\textrm BL}$ to represent the band limited version of the spin lowering operator, that is relevant in the case of a discretely sampled spin-2 field on the sphere. In the limit of $\ell_{\rm max} \rightarrow \infty$, the $\bar{\eth}^2_{\textrm BL} \rightarrow \bar{\eth}^2$.
The band limited version of the spin raising operator $\eth^2$ is derived by simply taking the conjugate of the above equation.
%--------------------------------------------------------

### 2018-07-31 articles/formalism/sections/kernel_paintbrush.tex
Added some clarifications based on Mathieu comments.

Changing the polarization direction within the filament changes the $E/B$ patterns.  A $90^\circ$ rotation of the polarization with respect to the filament changes the sign of both $E$ and $B$.  In a polarization pattern aligned at $45^\circ$ to the filament, the $E$ pattern will swap with the $B$ pattern.  The work in \cite{Zaldarriaga2001a} correctly argued that linear filaments (infinite and without ends) can only produce $E$-power, but we find that for realistic filaments, the ends produce $B$-mode power with a clear signature.  Careful study of the $E/B$ mode power in filaments can provide insights into the orientation of the magnetic field with respect to the axis of the filament, this information could potentially shed light on the internal dynamics of filaments.
% 
\begin{figure}[t]
\includegraphics[width=0.5\columnwidth]{line.pdf}
\includegraphics[width=0.5\columnwidth]{spiral.pdf}
\includegraphics[width=1.0\columnwidth]{colorbar.pdf}
\caption{ The polarization signals of toy filament structures. In a filament organized perfectly along a magnetic field line, the polarization will be perpendicular to the filament direction.  The $E/B$ modes of filaments are in some ways easier to think about than the Stokes parameters. Left panels: in a straight filament, the $E$-mode is positive along the filament and at the ends, but negative along the sides.  The $B$-modes are only non-zero at the ends.  Right panels: in a curved filament, the $E$-mode is again positive along the filament.  Outside the filament, the $E$-mode is more negative on the interior of the curve than the exterior.  The $B$-modes are again non-zero at the ends, akin to the straight filament case, but also somewhat along the filament due to the changing radius of curvature. In all images, the longitude angle increases to the left (which is East in sky convention).  All plot are on a common, arbitrary color scale.}
\label{fig:polfilaments}
\end{figure}
%
\indent The intuition from the real-space kernels holds also when we distort the shape of the filament.  If the filament were bent around into a circle, the positive and negative parts of the $B$ pattern will cancel, and we are left with a hoop of pure $E$ pattern. Here it is important to note that this cancellation will not happen for an ellipse or any loop of non-constant radius of curvature.
The same general description holds for a spiral-shaped filament, which can can be viewed as distortion of the straight filament (right panel of  \fig{fig:polfilaments}).  The filament is highlighted by positive $E>0$.  The $E$-pattern is more negative on the interior of a curve than on the exterior, and the concentric rings of filamentary structure make an increasingly negative $E$ value inside. While the $B$-pattern is again concentrated at the ends of the filament in an oriented pair of positive/negative fluctuations, note that it is non-vanishing in the intermediate regions along the filament due to the varying radius of curvature of the spiral.  A spiral turning with the other handedness will have a $B$-pattern with the opposite sign.

A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but claims no $B$-mode signal above the noise. We predict that a $B$-mode signal from filaments should be present, since they have a finite length of a few degrees.  Detecting $B$-modes from filaments will be easier with more signal-to-noise and may require a more careful filament analysis, rescaling and aligning the filament ends.  While the detectability of the $B$-mode signature from stacked filaments in Planck data calls for a more careful assessment, we should see it in higher-fidelity data.

>>>>

Changing the polarization direction within the filament changes the $E/B$ patterns.  A $90^\circ$ rotation of the polarization with respect to the filament changes the sign of both $E$ and $B$.  In a polarization pattern aligned at $45^\circ$ to the filament, the $E$ pattern will swap with the $B$ pattern.  The work in \cite{Zaldarriaga2001a} correctly argued that linear filaments (infinite and without ends) can only produce $E$-power, but we find that for realistic filaments, the ends produce $B$-mode power with a clear signature.  Careful study of the $E/B$ mode power in filaments can provide insights into the orientation of the magnetic field with respect to the axis of the filament, this information could potentially shed light on the internal dynamics of filaments.
% 
\begin{figure}[t]
\includegraphics[width=0.5\columnwidth]{line.pdf}
\includegraphics[width=0.5\columnwidth]{spiral.pdf}
\includegraphics[width=1.0\columnwidth]{colorbar.pdf}
\caption{ The polarization signals of toy filament structures. In a filament organized perfectly along a magnetic field line, the polarization will be perpendicular to the filament direction.  The $E/B$ modes of filaments are in some ways easier to think about than the Stokes parameters. Left panels: in a straight filament, the $E$-mode is positive along the filament and at the ends, but negative along the sides.  The $B$-modes are only non-zero at the ends.  Right panels: in a curved filament, the $E$-mode is again positive along the filament.  Outside the filament, the $E$-mode is more negative on the interior of the curve than the exterior.  The $B$-modes are again non-zero at the ends, akin to the straight filament case, but also somewhat along the filament due to the changing radius of curvature. In all images, the longitude angle increases to the left (which is East in sky convention).  All plot are on a common, arbitrary color scale.}
\label{fig:polfilaments}
\end{figure}
%
\indent The intuition from the real-space kernels holds also when we distort the shape of the filament.  If the filament were bent around into a circle, the positive and negative parts of the $B$ pattern will cancel, and we are left with a hoop of pure $E$ pattern. Here it is important to note that this cancellation will not happen for an ellipse or any loop of non-constant radius of curvature.
The same general description holds for a spiral-shaped filament, which can can be viewed as the distortion of the straight filament (right panel of  \fig{fig:polfilaments}).  The filament is highlighted by positive $E>0$.  The $E$-pattern is more negative on the interior of a curve than on the exterior, and the concentric rings of filamentary structure make an increasingly negative $E$ value inside. While the $B$-pattern is again concentrated at the ends of the filament in an oriented pair of positive/negative fluctuations, note that it is non-vanishing in the intermediate regions along the filament due to the varying radius of curvature of the spiral.  A spiral turning with the other handedness will have a $B$-pattern with the opposite sign.

A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but claims no $B$-mode signal above the noise. We predict that a $B$-mode signal from the filaments should be present, since they have a finite length of a few degrees.  Detecting $B$-modes from filaments will be easier with more signal-to-noise and may require a more careful filament analysis, rescaling and aligning the filament ends.  While the detectability of the $B$-mode signature from stacked filaments in Planck data calls for a more careful assessment, we should see it in higher-fidelity data.
